Phonon polaritonics in two-dimensional materials

### 2019-03-19 paper.tex
Some updates to intro (micro+macro sentences) & some revisions of the 2D->3D and 2D->1D discussion

@@

% --- ABSTRACT ---
\begin{abstract}
    Extreme confinement of electromagnetic energy by phonon polaritons holds the promise of strong and new forms of control over the dynamics of matter. 
    To bring such control to the atomic-scale limit, it is important to consider phonon polaritons in two-dimensional (2D) systems. Recent studies have pointed out that in 2D, splitting between longitudinal and transverse optical (LO and TO) phonons is absent at the $\Gamma$ point, even for polar materials.  
    Does this lack of LO--TO splitting imply the absence of a phonon polariton in polar monolayers? 
    Here, we derive a first-principles expression for the conductivity of a polar monolayer specified by the wavevector-dependent LO and TO phonon dispersions. In the long-wavelength (local) limit, we find a universal form for the conductivity in terms of the LO phonon frequency at the $\Gamma$ point,
    its lifetime, and the group velocity of the LO phonon. Our analysis reveals that the phonon polariton of 2D is simply the LO phonon of the 2D system. For the specific example of hexagonal boron nitride (hBN), we estimate the confinement and propagation losses of the LO phonons, finding that high confinement and reasonable propagation quality factors coincide in regions which may be difficult to detect with current near-field optical microscopy techniques. Finally, we study the interaction of external emitters with two-dimensional hBN nanostructures, finding extreme enhancement of spontaneous emission due to coupling with localized 2D phonon polaritons, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons. This may lead to the design of new hybrid states of electrons and phonons based on strong coupling.

@@

We now develop a framework with quantitative, concrete examples that enables a first-principles understanding of phonon polaritons in 2D materials. 
We do so by deriving a universal form for the conductivity of a polar monolayer,  which depends solely on the LO and TO phonon frequencies---and their dispersion with momentum---in the 2D system. Using parameters from Ref.~\citenum{sohier2017breakdown} for the canonical 2D polar monolayer---hexagonal boron nitride (hBN)---we present the confinement and propagation losses of the 2D phonon polariton modes, identifying the frequency region where they should be most easily detected. Finally, we find that these modes enable extreme light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced by up to eight orders of magnitude through the emitter--LO phonon coupling. For an infrared emitter with a sufficiently high free-space radiative decay rate (${\gtrsim}\, \SI{1e6}{\per\s}$), we find that the associated linewidth of the emitter is comparable to the spacing between different phonon polaritonic resonances of an hBN nanostructure. This suggests the possibility of realizing the multi-mode strong coupling and ultra-strong coupling regimes of quantum electrodynamics in a 2D hBN platform. Our results for hBN are particularly relevant due its widespread use in 2D van der Waals heterostructures. In addition to providing functionality as a layer which improves the electrical and optical properties of other 2D materials, \eg graphene, our results suggest that in these heterostructures, hBN layers could provide a mid-infrared platform for nanophotonics and quantum optics. While we focus on hBN in this manuscript, the salient features of our findings apply to other polar monolayers as well.

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO--TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}$''  should be considered equal to zero. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.

\addition{%
The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. 
One may describe the LO--TO splitting in terms of Eq.~\eqref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, the oscillator strength scales as $N$ \cite{sohier2017breakdown}. This should enhance the LO--TO splitting by a factor of $N$ and adjust the observed phonon polariton frequencies accordingly. To go beyond an effectively 2D description of a monolayer, and make the transition from monolayer to bulk, we may make use of the quasi-2.5D Coulomb interaction~\cite{thygesen2017calculating} in Eq.~\eqref{eq:zerosofresponse}. 
In that case, for a film of effective thickness $t$, the interaction is approximately $V_{\text{Q2.5D}}(q) = \frac{e^2}{\epsilon_0\epsilon_{\infty} tq^2}\big(1-\frac{2}{qt}e^{-\frac{qt}{2}}\sinh\frac{qt}{2} \big)$. 
This form interpolates between the 2D and 3D interactions in the limits $qt \rightarrow 0$ and $qt \rightarrow \infty$, assuming that $\epsilon_{\infty}$ also interpolates between the monolayer value, which is 1, and the bulk value, which is $\epsilon_{\infty}$. For sufficiently large thickness and finite-$q$, the 2.5D Coulomb interaction scales as $t^{-1}q^{-2}$. 
Noting the $q^2$ factor in Eq.~\eqref{eq:zerosofresponse}, as well as the $N$-dependence of the quantity involving eigendisplacements, one finds an LO--TO splitting independent of thickness, as expected for bulk samples.
}

\addition{%
For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. 
In that case, the 1D Coulomb interaction is $e^2K_0(qr)/2\pi\epsilon_0$, where $K_0$ is the (zeroth order) modified Bessel function of the second kind. 
Thus, the $q$-dependence of the LO--TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the discussion of Figure 1. 
As regards the optical properties, the optical conductivity is still governed by Eq.~\eqref{eq:2dsusceptibility}.
}

>>>>

\newcommand{\appropto}{\mathrel{\vcenter{
			\offinterlineskip\halign{\hfil$##$\cr
				\propto\cr\noalign{\kern.5pt}\sim\cr\noalign{\kern-1pt}}}}}

@@

% --- ABSTRACT ---
\begin{abstract}
    Extreme confinement of electromagnetic energy by phonon polaritons holds the promise of strong and new forms of control over the dynamics of matter. 
    To bring such control to the atomic-scale limit, it is important to consider phonon polaritons in two-dimensional (2D) systems. Recent studies have pointed out that in 2D, splitting between longitudinal and transverse optical (LO and TO) phonons is absent at the $\Gamma$ point, even for polar materials.  
    Does this lack of LO--TO splitting imply the absence of a phonon polariton in polar monolayers? 
	\addition{To answer this, we connect the microscopic phonon properties with the macroscopic electromagnetic response.}
	Specifically, we derive a first-principles expression for the conductivity of a polar monolayer specified by the wavevector-dependent LO and TO phonon dispersions. In the long-wavelength (local) limit, we find a universal form for the conductivity in terms of the LO phonon frequency at the $\Gamma$ point,
    its lifetime, and the group velocity of the LO phonon. Our analysis reveals that the phonon polariton of 2D is simply the LO phonon of the 2D system. For the specific example of hexagonal boron nitride (hBN), we estimate the confinement and propagation losses of the LO phonons, finding that high confinement and reasonable propagation quality factors coincide in regions which may be difficult to detect with current near-field optical microscopy techniques. Finally, we study the interaction of external emitters with two-dimensional hBN nanostructures, finding extreme enhancement of spontaneous emission due to coupling with localized 2D phonon polaritons, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons. This may lead to the design of new hybrid states of electrons and phonons based on strong coupling.

@@

\addition{%
To resolve this question, a connection between the microscopic phonon properties and the macroscopic electromagnetic response is required.
}
To that end, we develop a first-principles framework for phonon polaritons in polar 2D materials and illustrate it with quantitative, concrete examples.
We derive a universal form for the conductivity of a polar monolayer,  which depends solely on the LO and TO phonon frequencies---and their dispersion with momentum---in the 2D system. Using parameters from Ref.~\citenum{sohier2017breakdown} for the canonical 2D polar monolayer---hexagonal boron nitride (hBN)---we present the confinement and propagation losses of the 2D phonon polariton modes, identifying the frequency region where they should be most easily detected. Finally, we find that these modes enable extreme light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced by up to eight orders of magnitude through the emitter--LO phonon coupling. For an infrared emitter with a sufficiently high free-space radiative decay rate (${\gtrsim}\, \SI{1e6}{\per\s}$), we find that the associated linewidth of the emitter is comparable to the spacing between different phonon polaritonic resonances of an hBN nanostructure. This suggests the possibility of realizing the multi-mode strong coupling and ultra-strong coupling regimes of quantum electrodynamics in a 2D hBN platform. Our results for hBN are particularly relevant due its widespread use in 2D van der Waals heterostructures. In addition to providing functionality as a layer which improves the electrical and optical properties of other 2D materials, \eg graphene, our results suggest that in these heterostructures, hBN layers could provide a mid-infrared platform for nanophotonics and quantum optics. While we focus on hBN in this manuscript, the salient features of our findings apply to other polar monolayers as well.

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\big|^2 \equiv Q_*^2/M_*$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO--TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}$''  should be considered equal to zero. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.

\addition{%
The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which remain two-dimensional from an electromagnetic perspective for polariton wavelengths large compared to the bilayer thickness. 
Concretely, the LO--TO splitting in such quasi-2D systems is still given by Eq.~\eqref{eq:zerosofresponse}, provided the appropriate eigendisplacements of the few-layer system are employed, and provided suitable modifications of the Coulomb interaction are made.
For $N$ weakly coupled layers, the sum of Born effective charges $\big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\big|^2$, which is essentially an oscillator strength, is approximately $NQ_*^2/M_*$, \ie enhanced $N$-fold relative to the monolayer or, equivalently, proportional to the thickness $t$~\cite{sohier2017breakdown}. 
For few layers, \ie for $qt\ll 1$ where the Coulomb interaction remains essentially 2D, this simple geometric effect enhances the LO--TO splitting by a factor $N$.
At larger $qt$, the Coulomb interaction must eventually approach its 3D limit; to study this transition from monolayer to bulk explicitly, a quasi-2.5D Coulomb interaction~\cite{thygesen2017calculating} can be employed in Eq.~\eqref{eq:zerosofresponse}. 
Specifically, for a film of thickness $t$, the interaction is approximately $V_{\text{Q2.5D}}(q) = \frac{e^2}{\epsilon_0\epsilon_{\infty} q^2t}\big(1-\frac{2}{qt}e^{-\frac{qt}{2}}\sinh\frac{qt}{2} \big)$, with the static electronic screening $\epsilon_{\infty}$ interpolating between its monolayer value, 1, and its bulk value, $\epsilon_{\infty}$. 
The $qt \rightarrow 0$ and $qt \rightarrow \infty$ asymptotics of this interaction are the 2D and 3D interactions---$e^2/2\epsilon_0q$ and $e^2/\epsilon_0\epsilon_\infty q^2t$, respectively---which it interpolates between. 
In the large-thickness or -momentum limit, $qt\gg 1$, the ${\appropto}\,t$ scaling of the oscillator strength, the ${\appropto}\,t^{-1}q^{-2}$ scaling of the Coulomb interaction, and the ${\propto}\,q^2$ factor in Eq.~\eqref{eq:zerosofresponse} cancel, producing a thickness- and momentum-independent LO--TO splitting, as expected in the bulk. 
}

\addition{%
Analogously, our considerations can be extended to 1D polar materials, such as BN nanotubes, by employing the 1D Coulomb interaction (and, naturally, the 1D-specific eigendisplacements).
For a 1D system of transverse extent $R$ (\eg, the radius of a nanotube), the 1D Coulomb interaction is $e^2K_0(qr)/2\pi\epsilon_0$, where $K_0$ is the (zeroth order) modified Bessel function of the second kind. 
Thus, the $q$-dependence of the (squared) LO--TO splitting is $q^2K_0(qR)\simeq - q^2 \log q R$. 
This closely mirrors the situation in 1D plasmonics, as expected from the discussion of Figure 1. 
%As regards the optical properties, the optical conductivity is still governed by Eq.~\eqref{eq:2dsusceptibility}.
}
